---
title: "Frequently Asked Questions"
author: "Matti Vuorre"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    df_print: kable
bibliography: bibliography.bib
---

```{r setup, include = FALSE}
library(knitr)
library(bmlm)
opts_chunk$set(message = F, warning = F)
```

# Subject-specific parameters and effects

## How to identify specific subjects' effects?

When plotting subject specific effects using `mlm_pars_plot()` or `mlm_spaghetti_plot()`, individual subjects' effects or fitted values are represented as points or lines. Sometimes it might be of interest to identify specific subjects with lines / points in the figures.

One solution to identifying specific subjects in the estimated model is to inspect the numerical values of the estimated parameters: All subject-specific effects' posterior samples are saved in the estimated model object as *u_x*, where *x* is the parameter (one of *a*, *b*, *cp* (c'), *c*, *me*, or *pme*). 

For example, let's focus on the subject-specific *a* parameters. Let's identify the 6 subjects with the lowest values for this parameter. First, we'll fit a model using the included `BLch9` data set:

```{r subject-effects-estimate, cache = T}
library(bmlm)
head(BLch9)
fit <- mlm(BLch9, x="x", m="m", y="y", cores = 4, iter = 500)
```

Once the model has been estimated, you can find the *u_a* parameters with `mlm_summary(fit, pars = "u_a")`. However, this would return a row of information for each subject in the data. What we would like to do here is to arrange the estimated *u_a* parameters in an (ascending) order of their magnitudes (calculated from posterior means). To do that, we'll use the `arrange()` function [@wickham_dplyr:_2016] to arrange the data frame returned by `mlm_summary()`

```{r subject-effects-2}
library(dplyr)
u_a <- mlm_summary(fit, pars = "u_a")
head( arrange(u_a, Mean) )
```

Notice that we used the `head()` function to return the first six rows of the resulting data frame. The subject numbers are in square brackets: ID `r substr(arrange(u_a, Mean)[1,1], 5, 6)` has the lowest *u_a* parameter value. To find the individuals with the highest *u_a* values, wrap `Mean` in `desc()` (this arranges the data frame on descending values of Mean).

```{r}
head( arrange(u_a, desc(Mean)) )
```

**Important**: To use this method accurately, please ensure that the subjects' ID numbers in your data are sequential integers from 1 to however many subjects there are in your sample. The underlying Stan MCMC sampler requires sequential integer subject IDs, and if your IDs are something else (letters, non-sequential integers, etc.) `mlm()` will coerce the subject numbers to 1:N sequential integers before fitting the model.

Thanks to Xiao Hu for asking this question.

# References
