---
title: "bmlm: Multilevel Mediation Modeling with Stan"
author: "Matti Vuorre"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{bmlm: Multilevel mediation modeling with Stan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

[__bmlm__](https://github.com/mvuorre/bmlm) is an R package that allows easy estimation of multilevel mediation models. It relies heavily on the [RStan](http://mc-stan.org/interfaces/rstan.html) interface to the powerful  [Stan](http://mc-stan.org/) Bayesian inference engine. Users can estimate, summarize and plot a multilevel mediation model easily with the convenient functions provided with __bmlm__. This document explains how to install __bmlm__ and its required components, and then walks through an example of how to use it in practice.

# Installing __bmlm__

## Pre-requisites 

1. Please ensure you have the [latest version of R](https://cran.r-project.org/) installed.
2. This package depends on [Stan](http://mc-stan.org/). Please see [here](http://mc-stan.org/interfaces/rstan.html) for how to install the RStan interface.
3. Currently, bmlm is only available on GitHub, and therefore requires [devtools](https://cran.r-project.org/web/packages/devtools/index.html) for installation. If you don't have the devtools package installed in R, install it by running the following code in the R console:

```{r, eval = F}
install.packages("devtools")
```

## Installing bmlm

Then proceed to install bmlm:

```{r, eval = F}
devtools::install_github("mvuorre/bmlm")
```

If something goes wrong during the installation process, you will receive a notice usually asking you to install additional packages. If you are unable to resolve the problems, please contact me.

# Example

```{r, echo = F, message=FALSE, results='hide'}
# Loads pre-fitted model to save time building the vignette
load("fit.rda")
library(knitr)
opts_chunk$set(message = F, warning = F)
```

After installing the required software, load the __bmlm__ package to your current R workspace:

```{r, message=FALSE}
library(bmlm)
```

__bmlm__ contains an example data set from [Intensive Longitudinal Methods: An Introduction to Diary and Experience Sampling Research](http://www.intensivelongitudinal.com/index.html) (Bolger & Laurenceau, 2013). We'll use this data set in the example, and first load it into the workspace from the package, and display what the data looks like:

```{r, eval = F, echo = T}
data(BLch9)
head(BLch9)
```

```{r, eval = T, echo = F}
# Hack to show tabulated output instead of raw print
data(BLch9)
kable(head(BLch9), digits = 2)
```

## Data preprocessing

The goal of multilevel mediation modeling, in this case, is to assess the _within-person_ relationships between X, M and Y. To this end, it is important to isolate the within- and between-person components of X, M, and Y. The `isolate()` function in __bmlm__ allows the user to create within- and between-person centered values of variables, and additionally to standardize them. The example dataset `BLch9` already contains subject-mean deviated components, but here we illustrate how to obtain these using the `isolate()` function. The key inputs to this function are `d`, a data frame; `by` a column of values that identifies individuals; and `value`, which variable(s) should be transformed.

```{r}
BLch9 <- isolate(BLch9, 
                 by = "id", 
                 value = c("fwkstrs", "fwkdis", "freldis"))
```

```{r, echo = F}
kable(head(BLch9), digits = 2)
```

The `..._cw` variables now contain isolated within-person ("subject-mean deviated") pieces of each variable. We'll use these for the mediation analysis.

## Fit model

To estimate the multilevel mediation model, run `mlm()` and save its output to an object. Here we'll call it `fit`. You can also ask Stan to run multiple MCMC chains in parallel (if supported by your computer), as shown by the two first lines below.

```{r, eval = F}
rstan_options(auto_write = TRUE)  # These two lines are optional
options(mc.cores = parallel::detectCores()) 

fit <- mlm(d = BLch9, 
           id = "id",
           x = "fwkstrs_cw",
           m = "fwkdis_cw",
           y = "freldis_cw",
           iter = 4000)
```

The main arguments to `mlm()` are `d` (a `data.frame`), which here was set to `BLch9`. The user also needs to specify which columns contain the variables needed for the mediation model, unless they are already named `id`, `x`, `m`, and `y`:

* `id` is a column of participant IDs in the provided `data.frame`.
* `x` is the manipulated variable.
* `m` is the mediator variable.
* `y` is the outcome variable.

There are various additional arguments to the above command. Most notably, the `iter = 4000` specified that Stan should draw 4000 samples from the posterior distribution, for each MCMC chain. The default is to use 4 chains. Further, Stan's MCMC algorithms use a portion of the samples as warmup to adjust various underlying parameters. The default of one half was used for this example. 

Stan's MCMC procedures are very efficient, but estimating the model with large datasets will take a while. This example model (2100 rows of data) takes about 45 seconds on a desktop Mac (8GB RAM, 4ghz Intel i7).

## Summarize fitted model

After the samples have been obtained, __bmlm__'s helper functions can be used to obtain summaries of the results. For more options, all [rstan](https://cran.r-project.org/web/packages/rstan/index.html) methods are also available.

### Numerical summary

A numerical summary can be obtained by `mlm_summary(fit)`:

```{r, eval = F, echo = T}
mlm_summary(fit)
```

```{r, eval = T, echo = F}
# Hack to show tabulated output instead of raw print
kable(mlm_summary(fit))
```

`mlm_summary()` returns, for each parameter, the following information:

* Posterior _Mean_: This can be used as a point estimate of parameter
* Posterior _SD_: The standard deviation of the marginal posterior distribution of plausible parameter values. 
* Posterior _Median_: If the posterior distribution is skewed, the median might be used as a more accurate point estimate.
* Lower and upper limits to _Credible Intervals_. The CIs summarize the central X% mass of the marginal posterior distribution, where X% is defined by a `level` argument to `mlm_summary()`. The default "confidence level" is 0.91, but users may supply any value they desire.
* Posterior probability (_pprob_), which summarizes the probability density at the observed side of 0. For example, if the estimate (_Mean_) is -0.55, the posterior probability is defined as the proportion of probability density _below zero_.
* n_eff and Rhat are diagnostic values used to diagnose the performance of the underlying Stan MCMC procedures. see `?mlm` or `?stan` for details.

### Graphical summaries

Currently, __bmlm__ provides two graphical summaries of the estimated model. The first draws a path diagram of the mediation model, with point estimates of the relevant average-level parameters, and their associated credible intervals (as defined by `level`):

```{r, fig.width = 6, fig.height = 3.5, fig.cap="Path diagram of the average level mediation model, with numerical summaries of the relevant parameters."}
mlm_path_plot(fit, level = .89)
```

This figure offers a quick view of the estimated model: All paths (`a, b, c'`) are positive with fairly narrow credible intervals. The upper left corner gives three more parameters: `ab` is the average mediated effect, `c` is the total effect, and `%me` is the percent mediated effect. The width of each arrow represents the magnitude of each relationship, and the color indicates if the relationship is positive (green) or negative (red). The latter option can be turned off by passing argument `edge.color = "black"`.

For a more detailed investigation of the model's parameters, __bmlm__ offers two methods for plotting the samples from each parameter's posterior distribution. These are obtained by a call to `mlm_pars_plot(type = X)` where `X` can be either `hist` for histograms, or `coefplot` for a coefficient plot. If plotting a coefficient plot, the user may specify `level` to set the "confidence level", which is represented by the length of the lines surrounding each point estimate.

```{r, fig.cap = "Figure 2. A coefficient plot.", fig.show='hold'}
mlm_pars_plot(fit, type = "coefplot", level = .50)
mlm_pars_plot(fit, type = "coefplot", level = .99)
```

The user can also specify the parameters to show on the plot, as illustrated in the next histograms:

```{r, fig.height = 4, fig.width = 4, fig.cap = "Histograms for a selection of parameters."}
mlm_pars_plot(fit, type = "hist", pars = c("ab", "c", "pme", "corrab"))
```

The histograms are useful for visually assessing the shape of each marginal posterior distribution.

More complex plots are also possible:

```{r, fig.height = 5, fig.width = 5, fig.cap = "Varying effects standard deviations (tau) and correlations (Omega)."}
mlm_pars_plot(fit, type = "hist", pars = c(
    "tau[1]", "Omega[1,2]", "Omega[1,3]",
    "Omega[2,1]", "tau[2]", "Omega[2,3]",
    "Omega[3,1]", "Omega[3,2]", "tau[3]"),
    nrow = 3)
```

The first three positions (1: path `c'`, 2: path `b`, 3: path `a`) of the varying effects SDs ($\tau$) and correlations ($\Omega$) are plotted as histograms. Each histogram represents the MCMC samples of plausible parameter values from the corresponding posterior distribution.

## Person-specific effects

It is also possible to investigate person-specific parameters. These exist with the same name as the average-level parameters, but have "u_" appended to them. 

```{r, fig.width = 4, fig.height = 5, fig.cap = "Participant-specific c' path values with 67% Credible Intervals."}
head(mlm_summary(fit, pars = "u_c"))
mlm_pars_plot(fit, pars = "u_cp", type = "coef", level = .67) + 
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank())
```

# Further information

The functions contained in __bmlm__ have various options available for the user. Inspect these options by looking at the functions' help pages:

```{r, eval = F}
?mlm
?mlm_summary
?mlm_path_plot
?mlm_pars_plot
```

Users can also input any valid `stan()` arguments to `mlm()`; see `?stan` for details. The fitted object from `mlm()` is a valid `Stanfit` object, so all `Stanfit` methods are available as well. For example, it is important to assess the performance of Stan's MCMC algorithms after estimating a model by, for example, inspecting traceplots for convergence problems:

```{r, fig.width = 6}
pars <- c("a", "ab", "cp", "c", "corrab")
traceplot(fit, pars = pars)
```

Further modification of the model is possible through modifying the Stan code underlying `mlm()`. To obtain the Stan code for the multilevel mediation model, run `get_stancode()`

```{r}
cat(get_stancode(fit))  # Wrap inside cat() for nicer printing
```

