---
title: "bmlm: Multilevel mediation modeling with Stan"
author: "Matti Vuorre"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{bmlm: Multilevel mediation modeling with Stan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

[__bmlm__](https://github.com/mvuorre/bmlm) is an R package that allows easy estimation of multilevel mediation models. It relies heavily on the [RStan](http://mc-stan.org/interfaces/rstan.html) interface to the powerful  [Stan](http://mc-stan.org/) Bayesian inference engine. Users can estimate, summarize and plot a multilevel mediation model easily with the convenient functions provided with __bmlm__. This document explains how to install __bmlm__ and its required components, and then walks through an example of how to use it in practice.

# Installing __bmlm__

## Pre-requisites 

1. Please ensure you have the [latest version of R](https://cran.r-project.org/) installed.
2. This package depends on [Stan](http://mc-stan.org/). Please see [here](http://mc-stan.org/interfaces/rstan.html) for how to install the RStan interface.
3. Currently, bmlm is only available on GitHub, and therefore requires [devtools](https://cran.r-project.org/web/packages/devtools/index.html) for installation. If you don't have the devtools package installed in R, install it by running the following code in the R console:

```{r, eval = F}
install.packages("devtools")
```

## Installing bmlm

Then proceed to install bmlm:

```{r, eval = F}
devtools::install_github("mvuorre/bmlm")
```

If something goes wrong during the installation process, you will receive a notice usually asking you to install additional packages. If you are unable to resolve the problems, please contact me.
